---
title: "geoaddSAE2-intro"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{geoaddSAE2-intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", 
                      warning = FALSE, message = FALSE)
```

## Background

`geosae` fits the **area-level Geoadditive Small Area Estimation (Geoadditive
SAE)** model: a semiparametric extension of the Fay-Herriot model in which

* linear covariate effects are modelled as usual,
* nonlinear covariate effects are modelled with penalized splines (P-splines), and
* spatial variation is modelled with a bivariate thin plate regression spline,

all represented jointly as a linear mixed model and fitted by Restricted
Maximum Likelihood (REML). The Mean Squared Error (MSE) of the resulting
small area predictor is obtained by parametric bootstrap.

The classical Fay-Herriot (FH) and Spatial Fay-Herriot (SFH) models are
**not** re-implemented: `geosae` calls the existing implementations by the `sae` package so that the three approaches can be fitted and compared
consistently.

## A minimal example

```{r setup}
library(geoaddSAE2)
data(simulated_sae)
head(simulated_sae)
```

`simulated_sae` has a direct estimator `y`, a known sampling variance
`vardir`, a linear covariate `x1`, a nonlinear covariate `x2`, and spatial
coordinates `lat`/`lon`.

### Geoadditive SAE only

```{r}
fit <- geosae(
  data      = simulated_sae,
  formula   = y ~ x1,
  vardir    = vardir,
  nonlinear = "x2",
  spatial   = c("lat", "lon"),
  bootstrap = TRUE,
  B         = 50,
  seed      = 1
)

fit
```

```{r}
summary(fit)
```

### Comparing Geoadditive SAE, Fay-Herriot, and Spatial Fay-Herriot

Set `compare = TRUE` to also fit FH (always) and SFH (whenever `spatial` is
supplied), and get a side-by-side comparison table.

```{r}
fit_cmp <- geosae(
  data      = simulated_sae,
  formula   = y ~ x1,
  vardir    = vardir,
  nonlinear = "x2",
  spatial   = c("lat", "lon"),
  compare   = TRUE,
  B         = 50,
  seed      = 1
)
fit_cmp$comparison
```

## Inspecting results

A fitted `geosae` object cleanly separates:

* `fit$estimation` -- direct estimate, geoadditive estimate, MSE, RMSE per area;
* `fit$diagnostics` -- convergence, EDF of smooth terms, variance components, R-sq, deviance explained;
* `fit$parameters` -- fixed-effect (linear) coefficient table;
* `fit$comparison` -- model | mse | rmse | mape, when `compare = TRUE`;
* `fit$models` -- the underlying fitted model objects, for advanced use.

```{r}
# plot(fit, which = "smooth")
```
